---
title: "CMSC320 Final Project"
output: 
  html_document:
    number_sections: true
    toc: true


---


## Background
Graduate School is a goal for many students in University. However, unlike in high school, the resources and information are not as known, so prospective young adults have to independently do their research. There are many factors in the application process, and the purpose of this project is to help gain a better understanding on what is needed for a successful higher education.

## Setting up
Prior to accessing our dataset, we need to first load the packages necessary for this study.
```{r setupdb, include=TRUE, comment=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(broom)
library(caret)
library(plotROC)
library(magrittr)
library(dplyr)
library(ggplot2)
library(ggrepel)
```


# Data Curation

The dataset we will be using comes from a comma-separated values (CSV) file, found on https://www.kaggle.com/mohansacharya/graduate-admissions/downloads/graduate-admissions.zip/2
(need to cite properly). 
Mohan S Acharya, Asfia Armaan, Aneeta S Antony : A Comparison of Regression Models for Prediction of Graduate Admissions, IEEE International Conference on Computational Intelligence in Data Science 2019
^ on kaggle page

We can view the data by calling the read_csv command on the downloaded file.
```{r prep, comment=FALSE, warning=FALSE, message=FALSE}

csv<- "/Users/Maya/Downloads/Admission_Predict_Data.csv"
df <- read_csv(csv)
df

```

# Parsing and Management
Now that we have the data, we should understand what the components are.

## Entries
The entries for this dataframe are individuals who have applied to graduate school. Each entry, or row, corresponds to one person. 

## Attributes
The attributes (or columns) in the dataset are as follows: Serial No., GRE Score, TOEFL Score, University Rating, SOP, LOR, GPA, Research, and Chance of Admit

### Serial Number
The serial number of each entry merely identifies the row the entry is in. As a result, it does not need to be kept in the dataframe.
```{r remsno}
# remove first column containing Serial Number
tidy_data <- df[,-1]
tidy_data
```

### GRE Score
The GRE is the Graduate Record Examination, and serves as an admission test for graduate school. The exam is a broad assessment of one's critical thinking, analytica writing, verbal reasoning, and quantitative reasoning skills. The exam is scored from 130-170 in each of the two sections, Verbal and Quantitative, making the maximum possible score 340.

### TOEFL Score
The TOEFL is the test of English as a foreign language, taken by many international students. The test is scored on a scale of 0 to 120 and is made up of four sections: Reading, Listening, Speaking, and Writing. 

### University Rating
This is a rating out of 5 of the applicant's undergraduate university.

### Statement of Purpose (SOP)
The SOP is an essay written in the application, scored in this case from 1-5 in terms of strength.

### Letters of Recommendation (LOR)
The LOR are written by professors and mentors, and are scored in the dataset from 1-5 in terms of strength

### Cumulative GPA
The cumulative undergraduate GPA of the applicant is out of 10.0

### Research
The research attribute identifies whether or not an applicant conducted research prior to applying, with 1 for yes and 0 for no.
Since this is binary, we can turn the values into booleans, with true = 1 and false = 0

```{r research}
#convert 1's and 0's in Research to boolean values
tidy_data$Research <- as.logical(as.integer(tidy_data$Research))
tidy_data

```

### Chance of Admit
The chance of admittance into an American Graduate Program is scored from 0 to 1.

# Exploratory Data Analysis

# Hypothesis Testing


# Machine Learning
As students, we want to see what can maximize our chances of getting into a Graduate Program. Given that there are multiple factors in the dataset that may improve chances, we can conduct a linear regression to see if there is a relationship between any of the attributes.

## Simple Regression
To look at the relationship between two continuous variables, we can assume that in our population, the relationship between them is given by the linear function
$Y = \beta{_0} + \beta{_1}X \\$

Since we are interested in how chance of being admitted is affected by factors (ex: GRE score), we can do the following regression: 
$Chance of Admit \approx \beta{_0} + \beta{_1}\times GRE Score \\$

```{r simple}
simple_reg <- tidy_data %>%
  ggplot(aes(x=`GRE Score`, y= `Chance of Admit`)) +
  geom_point() +
  geom_smooth(method=lm)
simple_reg

# The linear model equation
gre_fit <- lm(`Chance of Admit`~`GRE Score`, data=tidy_data)
```
According to the lm function, $\hat\beta{_0} = -2.48281\\$ and $\hat\beta{_1} = 0.01013\\$ 
This means that on average, an applicant's chance of admittance would increase by 0.01 for every additional point scored on the GRE.

Now that we have a linear model, we must interpret how well it fits the data.
First, we need to check that there is a relationship between Chance of Admit and GRE Score. This can be done with a hypothesis test. 
$\ H{_0}; \beta{_1} = 0\\$ meaning there is no relationship
$\ H{_0}; \beta{_1} \neq 0\\$

```{r htest}
gre_fit %>%
  tidy()
```
We find that our p value for $ \hat\beta{_1} \\$ is much lower than .05, our cutoff for significance. Thus we can state that there is a statistically significant relationship between GRE Score and Chance of Admittance.

## Multiple Regression
We know that GRE is a factor for admittance, but what if there are more? In particular, what if a combination of multiple factors gives us a better model for the data?

